---
title: "An investigation into female literacy and fertility rate in Southeast Asian countries"
author: "Jue Zhou"
date: "11/7/2019"
output: 
  github_document: 
    toc: true
    toc_depth: 3
  pdf_document: 
    fontsize: 12pt
    geometry: margin = 1in 
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 4, 
  fig.retina = 2
)
```

## Introduction

Gary Becker, the famous economist and sociologist at the University of Chicago, was renowed for his attribution to the study of human capital. He argued that people's decision on child bearing can be seen as rational and utility maximizing, and thus constructed an economic model for family's child bearing decisions. His theory of fertility suggests that women's educational level influence fertility rate in two ways: first is an incentive effect, that more educated women have higher opportunity costs of bearing children in terms of lost income; second is the change in household bargaining power, which assumes that more educated women are better able to support themselves and have more bargaining power, including on family size (Becker, 1960).

While follow-up research generally supports the effect of female literacy on fertility decline, some recent studies report no significant correlation between female education and fertility and attribute the changes to other factors, such as the rise of income, urbanization, "son preference", access to public health services. (e.g., United Nations, 1993; Jeffery and Jeffery, 1996; Vlassoff, 1996)

In this report, I intend to conduct an exploratory analysis into the relationship between fertility decline and female education using the cases of several Southeast Asian countries: Indonesia, Malaysia, Philippines, Thailand, and Viet Nam. They were named as *Tiger Cub Economies* in 1990s for their similar export-driven economic policy. At the same time of newly industrial transition in the 1990s, they also experienced a population transition marked by the decrease in fertility rate. Similar economic development model, cultural backgrounds and lack of family planning programmes make these countries comparable.

## Data

Data I used in my analysis comes from **Wittgenstein Centre for Demography and Global Human Capital** (Available at [this link](http://www.wittgensteincentre.org/dataexplorer).) Wittgenstein centre provides a set of different scenarios of future population and human capital trends in 201 countries of the world to 2100. The result and update of the population projections by levels of educational attainment are based on the work of a large team of researchers at the Wittgenstein Centre for Demography and Global Human Capital and at other institutions. New version of the datasets also includes the reconstruction of population by levels of educational attainment from 1950 to 2015 for 185 countries. 

```{r load}
# load useful packages
library(tidyverse)
library(here)
library(wesanderson)

```


```{r dataset,  include = FALSE}
# load datasets and recode for analysis

# female education distribution by broad age (20-39)
here("data","wcde_data_femaleeducation.csv")
education <- read_csv(file = here("data","wcde_data_femaleeducation.csv"), skip = 8) %>% 
  filter(Education != "Under 15") %>% #since the sameple here are above 15
  # recode the stage of education
  mutate(Education = recode(Education,
                          "No Education" = 1,
                          "Incomplete Primary" = 2,
                          "Primary" = 3,
                          "Lower Secondary" = 4,
                          "Upper Secondary" = 5,
                          "Post Secondary" = 6,
                          "Short Post Secondary" = 6,
                          "Bachelor" = 7,
                         	"Master and higher" = 8)) %>% 
  # calculate the percent of female whose educational level is higher than lower secondary
  group_by(Area, Year) %>% 
  mutate(Highereducation = ifelse(Education > 4, Distribution, 0)) %>% 
  summarise(Highereducation = sum(Highereducation))

# Mean Years of Schooling by Broad Age (20-39, female)
here("data","wcde_data_school_year.csv")
schoolyears <- read_csv(file = here("data","wcde_data_school_year.csv"), skip = 8) %>% 
  select(Area, Year, Years) %>% 
  rename("MeanYearsSchooling" = Years)

# sex ratio in the range of 0-4
here("data","wcde_data_sex_ratio.csv")
sexratio <- read_csv(file = here("data","wcde_data_sex_ratio.csv"), skip = 8) %>% 
  rename("Sexratio" = Ratio)

# education gap between male and female (by years)
here("data","wcde_data_gender_gap.csv")
gendergap <- read_csv(file = here("data","wcde_data_gender_gap.csv"), skip = 8) %>% 
  rename("Educationgap" = Years)

# total fertility rate
here("data","wcde_data_tfr.csv")
tfr <- read_csv(file = here("data","wcde_data_tfr.csv"), skip = 8) %>% 
  rename("TFR" = Rate) %>% 
  mutate(Year = rep(c(1950,1955,1960,1965,1970,1975,1980,1985,1990,1995,2000,2005,2010,2015), times = 5)) %>% 
  select(-Period)

# combine the datasets
literacy_tfr <- education %>% 
  left_join(schoolyears) %>% 
  left_join(gendergap) %>% 
  left_join(sexratio) %>% 
  left_join(tfr, by = c("Area","Year")) %>% 
  rename("Country" = "Area")

```

## The General Change of Fertility Rate
Total fertility rate (TFR) in simple terms refers to total number of children born or likely to be born to a woman in her life time if she were subject to the prevailing rate of age-specific fertility in the population (WHO). The following graph shows the general change of fertility rate in the five countries listed above. All of them have experienced a dramatic decrease in fertility rate over the past half century, from over 6 to around 2. Three of them, Thailand, Viet Nam, and Malaysia, have reached the replacement level of fertility (2.1 children per woman). Although the fertitlity rate of the other four countries have kept stable in recent years, the fertility rate of Philippines is still in a decreasing trend.

```{r  fertility, echo = FALSE}
literacy_tfr %>%
  ggplot() +
  geom_smooth(aes(x = Year, y = TFR, color =  Country), se = FALSE) +
  scale_color_manual(values = wes_palette("Darjeeling1", 5, type = "continuous")) +
  labs(
    x = "Year",
    y = "Total Fertility Rate",
    title = "Graph 1: Change of Total Fertility Rate",
    caption = "Source: Wittgenstein Centre for Demography and Global Human Capital"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

## Findings

### Fertility rate and the average schooling years of women of childbearing age
```{r schoolyears}
literacy_tfr %>%
  ggplot() +
  geom_smooth(aes(x = MeanYearsSchooling, y = TFR, color = Country), se = FALSE) +
  scale_color_manual(values = wes_palette("Darjeeling1", 5, type = "continuous")) +
  labs(
    x = "Mean Years of Schooling (female aged 20-39)",
    y = "Total Fertility Rate",
    title = "Graph 2: Relationship between Total Fertility Rate and Average Schooling Years",
    caption = "Source: Wittgenstein Centre for Demography and Global Human Capital"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

To determine the relationship between female literacy and fertility rate, here I used the average schooling years of women in childbearing age to measure the level of female literacy. Similar to previous study, the result shows there is a negetive correlation between women's education years and total fertility rate.

### Fertility rate and percent of women receiving relatively higher education

Although the graph above shows the relationship between female education and fertility, there
appears to be some lack of clarity as to the pathways through which it operates. 
Here I hope to distingish between (i) the incentive effect----the opportunity costs of bearing children, and (ii) the change in household bargaining power.
For the incentive effect, I examined the relationship between percent of women in childbearing age who have received at least upper secondary education and total fertility rate. If higher education generally means higher income and higher opportunity costs of bearing children, would there be a negative relationship between these two variables?

```{r higheredu}
literacy_tfr %>%
  ggplot() +
  geom_smooth(aes(x = Highereducation/100, y = TFR, color = Country), se = FALSE) +
  scale_color_manual(values = wes_palette("Darjeeling1", 5, type = "continuous")) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1))+
  labs(
    x = "Percent of women aged 20-39 whose education level is at least upper secondary ",
    y = "Total Fertility Rate",
    title = "Graph 3: Relationship between Total Fertility Rate and Higher Education",
    caption = "Source: Wittgenstein Centre for Demography and Global Human Capital"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

Within single country level, the answer seems to be yes. The higher percent of women whose educational level is at least upper secondary, the lower total fertility rate goes. However, if we compare the relationship across countries, Philippines appears to be a very different case. Although the percent of women receiving higher level education exceeds all the other countries, its fertility rate is rather high.

### Fertility rate and educational gap between males and females
Household bargaining power model claims women's education not only increases the opportunity cost of childbearing, it also strengthens their bargaining power on the family size. When considering the power structure in a family, the male partner's socio-economic status should also be taken into consideration. Here I use the educational gap between male and female to measure the bargaining power change.

```{r edugap}
literacy_tfr %>%
  ggplot() +
  geom_path(aes(x = Educationgap, y = TFR, color = Country)) +
  scale_color_manual(values = wes_palette("Darjeeling1", 5, type = "continuous")) +
  labs(
    x = "Education Gap (Male-Female)",
    y = "Total Fertility Rate",
    title = "Graph 4: Relationship between Total Fertility Rate and Education Gap",
    caption = "Source: Wittgenstein Centre for Demography and Global Human Capital"
  ) +
  theme(plot.title = element_text(face = "bold", size = 16))
```

Similar to what we find above, we can see a positive correlation between the education gap and total fertility rate, which means as the closing of educational gap is correlated with the decline in total fertility rate. However, the case of Philipines is still very different from the other four countries. Its total fertility rate remains relatively high although its education gap has gone to negative in recent years.

Although here we do not have further data to dig into this phenomena, the fact that Philippines is not only a commodity-export country but also a large labor-export country might helps to explain it. According to a source in 2015, the Philippine government issued employment contracts to at least 1.8 million temporary migrant workers, the majority of them are women (Mendoza, 2015). If the female in the family is aborad working as migrant workers, and the male takes on most part of the job of child rearing, the opportunity cost and household bargaining power no longer serves as efficient mechanisms for decline in fertility rate.

### Fertility rate and Sex Ratio

Besides the factor of female literacy, some scholars also argue for the possible role of ‘son preference’ in influencing fertility. This concern arises from the common-sense observation that desire for a specified number of sons often interferes with the transition towards small-family norms.

Here I used the sex ratio of the population aged 0-4 to measure the gender preference in childbearing.

```{r sexratio}
literacy_tfr %>%
  ggplot() +
  geom_boxplot(aes(x = Sexratio, y = TFR, color = Country)) +
  scale_color_manual(values = wes_palette("Darjeeling1", 5, type = "continuous")) +
  labs(
    x = "Sex Ratio (0-4)",
    y = "Total Fertility Rate",
    title = "Graph 5: Relationship between Total Fertility Rate and Sex Ratio",
    caption = "Source: Wittgenstein Centre for Demography and Global Human Capital"
  ) +
  theme(plot.title = element_text(face = "bold", size = 16))
```

The graph shows that except for Thailand, the higher sex ratio of 0-4 age range is (which might reflects a son preference), the higher total fertility rate is, which indicates son preference might also plays a role in enhancing fertility rate.

## Conclusion

Analysis above indicates that increase in women's educational level does have a negative influence on total fertility rate of a country. However, comparison across countries shows that female literacy is not the only mechanism. Instead, economic mode, division of labor and gender preference might also play a part in dermining the fertility rate. Obviously, simple correlation analysis is not enough to shed light on the causal relationships and mechanisms behind the population transition. I hope to further investigate into this problem by applying multivariate regression and other statistical methods in the following weeks.


## Bibliography

Becker, Gary S. 1960. "An Economic Analysis of Fertility." In National Bureau Committee for Economic Research, Demographic and Economic Change in Developed Countries, a Conference of the Universities. Princeton, N.J.: Princeton University Press

Jeffery, R and P Jeffery. 1997. "Population, Gender and Politics: Demographic Change
in Rural North India." Cambridge: Cambridge University Press.

Mendoza, D. R. 2015. "Human capital: The Philippines’ labor export model". World Politics Review.

United Nations. 1993, Women's Education and Fertility Behavior: A Case Study of
Rural Maharashtra. New York: Department of Economic and Social
Information and Policy Analysis, United Nations.

Vlassoff, Carol. 1996. "Against the Odds: The Changing Impact of Schooling on
Female Autonomy and Fertility in an Indian Village". In Jeffery, R and A Basu. 1996. Girls' Schooling, Women's Autonomy, and Fertility Change in South Asia. New Delhi: Sage.


```{r, include=FALSE}
devtools::session_info()
```